---
title: "Spatial multivariate methods"
subtitle: "in practice"  
author: "Stéphane Dray"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [custom.css, xaringan-themer.css]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE, width = 72)
knitr::opts_chunk$set(tidy=TRUE, tidy.opts=list(width.cutoff=50), fig.height= 4, fig.width=4, dev='svg', fig.align="center", fig.path = "fig/")
```

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
duo(primary_color = "#1F4257", secondary_color = "#F97B64")
```

```{r include = FALSE}
library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})
``` 
$$\newcommand{\tr}{\hspace{-0.05cm}^{\top}\hspace{-0.05cm}} % transpose d'une matrice$$

$$\newcommand{\mb}[1]{\mathbf{#1}}$$

$$\newcommand{\sqnorm}[2]{
 \left \|  #1  \right \|^2_{#2}
}$$

$$\newcommand{\norm}[2]{
 \left \| #1 \right \|_{#2}
}$$


---
## Spatial data in R

Several packages allows to deal with spatial objects in R.

* The package `sp` provides classes to manage spatial objects with or without data

* Spatial weighting matrices can be defined and managed with `spdep`

* Spatial multivariate methods are implemented in `adespatial`

* Spatial objects can be used as arguments in `adegraphics` functions  to create maps

```{r, message = FALSE}
library(ade4) ; library(adegraphics) ; library(adespatial) ; library(spdep)
```


---

## `adespatial`

A package for spatial multivariate analysis interfacing `ade4` and `spdep`

More details in the vignette:

https://cran.r-project.org/web/packages/adespatial/vignettes/tutorial.html

or

```{r, eval = FALSE}
vignette("tutorial", package = "adespatial")
```

---

## Spatial data

* Classes are provide in R to deal with spatial data (raster, polygons, lines, points)

* We will use implementations in `sp` package but it will be replaced by `sf` in the future

* Import/export functions allows to interface with GIS software

* Spatial proximities are managed by `spdep` functionalities

---

## Spatial weighting matrix

A two-step procedure:

* define a binary neighboring matrix (`nb` object)

* consider optional non-binary weights and standardization to define a spatial weighting matrix (`listw` object)

```{r, out.width="40%"}
data(mafragh)
lw <- nb2listw(mafragh$nb)
s.label(mafragh$xy, plabel.cex = 0, Sp = mafragh$Spatial.contour, nb = mafragh$nb)
```

---

## Spatial mapping

```{r}
s.value(mafragh$xy, mafragh$flo[,1:4], method="size", symbol = "circle", Sp = mafragh$Spatial.contour)
```

---

## Spatial autocorrelation

Compute Moran's index for the five first species:
```{r}
moran.randtest(mafragh$flo[,1:5], lw)
```


---

## Multispati
### Perform the analysis

```{r}
pca_veg <- dudi.pca(mafragh$flo, scale = FALSE, scannf = FALSE)
ms_veg <- multispati(pca_veg, lw, scannf = FALSE) #<<

```

---
## Have a look to the summary
```{r, output.lines = 15}
summary(ms_veg)
```
---

## Plot the results

```{r, fig.width = 3.5, fig.height=3.5}
g1 <- plot(ms_veg)
names(g1)
```
---
## Loadings for variables

```{r, fig.width=3, fig.height=3}
g1$loadings
```

  $\mb{A}^*$: coefficients (loadings) for the variables of $\mb{X}$ (`ms_veg$c1`)
---
## Scores for individuals and lagged scores

```{r, fig.width = 3.5, fig.height=3.5}
g1$row
```
mem_sub
* $\mathbf{XQ}\mb{A}$: scores of individuals (`ms_veg$li`)
* $\mathbf{WXQ}\mb{A}$: lagged scores (`ms_veg$ls`)

---

## Moran's Eigenvectors Maps

```{r}
me <- mem(lw)
s.value(mafragh$xy, me[,1:4], Sp = mafragh$Spatial.contour)
```

---
### Variable selection
  
```{r, cache=TRUE, }
mem_sub <- mem.select(mafragh$flo, lw)  
mem_sub$summary
```

---
### Spatial RDA

```{r}
rda_spat <- pcaiv(pca_veg, mem_sub$MEM.select, scannf = FALSE)
plot(rda_spat)
```

---

### Variation partitioning

```{r}
varipart(pca_veg$tab, mem_sub$MEM.select, mafragh$env)
```

---

## Your turn
1. Create a Rmd or a R file
2. Load the `irishdata` data set from `ade4`
3. See `?irishdata` for details
4. Perform a PCA on the data table `irishdata$tab`
5. Define a spatial weighting matrix from `irishdata$Spatial` using `poly2nb`
6. Compute Moran's index of autocorrelation for the PCA scores
7. Perform MULTISPATI analysis
6. Interpret
