---
title: "Correspondence Analysis"
subtitle: "In practice"  
author: "Stéphane Dray"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css:  [custom.css, xaringan-themer.css] 
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    includes:
            before_body: preamble.tex
params:
  correction: FALSE
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(tidy=TRUE, tidy.opts=list(width.cutoff=50), fig.height= 4, fig.width=4, dev='svg', fig.align="center", fig.path = "fig/")
```

```{r correction, include = FALSE}
res.correction <- ifelse(params$correction, "markup", "hide")
```

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_duo(primary_color = "#1F4257", secondary_color = "#F97B64", table_row_even_background_color = "#1F4257")
```

---
# Data

We consider the relationship between hair and eyes colors measured on 592 students in Delaware University.

```{r}
mydata <- read.table("http://pbil.univ-lyon1.fr/R/donnees/snee74.txt", header=TRUE, stringsAsFactors = TRUE)
names(mydata)
head(mydata)
```

---
# Factors in R

A qualitative variables is stored as `factor` with different categories (`levels`)

```{r}
is.factor(mydata$yeux)
levels(mydata$yeux)
```

---
# Univariate analysis

* Compute the vector with the number of individuals for each eye color category and display the results on a plot.

```{r, echo = params$correction, include = params$correction, results=res.correction, out.width="30%"}
summary(mydata$yeux)
plot(mydata$yeux)
```

---
# Contingency table

* Build the contingency table crossing hair and eyes colors. See `?table`

```{r, echo = params$correction, results=res.correction}
mytab <- table(mydata$cheveux, mydata$yeux)
```

* Compute row and columns marginal relative frequencies
```{r, echo = params$correction, results=res.correction}
rowSums(mytab) / sum(mytab)
colSums(mytab) / sum(mytab)
```
---
# Bivariate analysis

* Compute and interpret the results of a $\chi^2$ test
```{r, echo = params$correction, results=res.correction}
chisq.test(mytab)
```
---
# Bivariate analysis

* Display and interpret the associations between categories with `mosaicplot`

```{r, echo = params$correction, include = params$correction, results=res.correction, out.width="50%"}
mosaicplot(mytab, shade = TRUE)
```

---
# Scoring

Suppose that we can display hair colors by a score

```{r, echo=FALSE, fig.width = 6, fig.height = 2, out.width = "70%"}
library(adegraphics)
set.seed(12)
score.h <- rnorm(4)
s1d.label(score.h, levels(mydata$cheveux))
```


what would be the best way to display eyes colors?
(given the distribution of students)
```{r, echo=FALSE, fig.width = 6, fig.height = 2, out.width = "70%"}
score.e <- rnorm(4)
s1d.label(score.e, levels(mydata$yeux))
```

---
# Weighted averaging

* Assign a random score to hair colors

```{r, echo = params$correction, results=res.correction}
set.seed(12)
score.h <- rnorm(4)
```

* Compute the position for the eye color 'Vert'

```{r, echo = params$correction, results=res.correction}
sum(score.h * mytab[,4] / sum(mytab[,4]))

## or

weighted.mean(score.h, w = mytab[,4])
```


---
# Compute scores for all eye colors

By repeating the same formulas
```{r}
score.e <- sapply(1:4, function(x) sum(score.h * mytab[,x] / sum(mytab[,x])))
```


or using matrix algebra

```{r}
t(prop.table(mytab, 2)) %*% score.h
```

---
# Best representation

if 
```{r, echo=FALSE, fig.width = 6, fig.height = 2, out.width = "70%"}
library(adegraphics)
s1d.label(score.h, levels(mydata$cheveux))
```

then 

```{r, echo=FALSE, fig.width = 6, fig.height = 2, out.width = "70%"}
s1d.label(score.e, levels(mydata$yeux))
```
---
# Reciprocal representation

* Start from a random score for eyes colors to position hair colors by averaging

```{r, echo = params$correction, results=res.correction}
set.seed(13)
score.e <- rnorm(4)
prop.table(mytab, 1) %*% score.e
```

---
# Reciprocal averaging

1. Set a random score for columns
2. Use the column score to compute a score for rows by weighted averaging
3. Compute a new score for columns by weighted averaging of row score
4. Center and scale the two scores using row and column weights
5. Repeat steps 2-4 until convergence

---
# Correspondence analysis

The solution of the iterative algorithm can also be obtained by the diagonalization procedure provided by the `dudi.coa` function


```{r, message = FALSE}
library(ade4)
coa1 <- dudi.coa(unclass(mytab), scannf = FALSE)
```

* Check the link between $\chi^2$ statistic and total inertia

```{r, echo = params$correction, results=res.correction}
sum(coa1$eig)
chisq.test(mytab)$statistic / sum(mytab)
```

---

# Interpretation

* Compute a row score with unit norm (`$l1`) to obtain a column score by weighted averaging (`$co`) with maximal variance (`$eig`)

* Compute a row score with unit norm (`$c1`) to obtain a column score by weighted averaging (`$li`) with maximal variance (`$eig`)

* Check both results

```{r, echo = params$correction, results=res.correction}
prop.table(mytab, 1) %*% coa1$c1[,1]
coa1$li[,1]
```

---
# Graphical representation

* Represent and interpret the results using the `scatter` function (check the `method` argument of `scatter.coa`)

```{r, echo = params$correction, include = params$correction, results=res.correction, out.width="50%"}
scatter(coa1, method = 2)
```

---
# Inertia statistics

* Use the `inertia.dudi` function to compute inertia statistics for rows and columns

```{r, echo = params$correction, results=res.correction}
ic = inertia.dudi(coa1, row.inertia=TRUE, col.inertia=TRUE)
```

* Represent absolute contributions using the `plot` function (see `?plot.inertia`)

```{r, echo = params$correction, include = params$correction, results=res.correction, out.width="60%", fig.width=6, fig.height=3}
plot(ic, contrib = "abs")
```
